---
title: "Iterative WGCNA Cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)

source("../utils/plot_theme.R")
source("../utils/helpers.R")
```

```{r load_data}
# Get cleaned data for covariates
cleaned <- readRDS("clean_data/1_rosmap_cleaning/rosmap_cleaned.RDS")

# Get submodule labels to convert back into module labels
modules <- read.table("raw_data/Modules_March2018.txt", sep="\t", header=T)
module.names <- as.character(unique(modules$Module))
module.names <- module.names[grep("DLPFC", module.names)]
```

I will read in specific files that were generated as output from `iterativeWGCNA`. This includes the final membership of genes (i.e. which submodule each gene belongs to after the iterative process) and eigengenes for each submodule. 

I save the cleaned files I have generated from the output from `iterativeWGCNA`. These are similar to the submodule data sets that Cai generated after using SVD decomposition on Ben's modules.

```{r}
# Create data frame for analyzing breakdown
breakdown <- list()

gene.set <- list()
eigengene.set <- list()
submodule.labels <- list()
for (i in 1:length(module.names)) {
  
  # Retrieve final membership of genes
  final.membership <- read.table(file=paste0("results/2_iterative_WGCNA_setup/", module.names[i], ".tsv/final-membership.txt"), header=T, sep="\t")
  
  # Check if file exists
  if (!file.exists(paste0("results/2_iterative_WGCNA_setup/", module.names[i], ".tsv/merged-0.05-eigengenes.txt"))) {
    next
  }
  
  # Retrieve eigengenes of submodules
  eigengenes <- read.table(file=paste0("results/2_iterative_WGCNA_setup/", module.names[i], ".tsv/merged-0.05-eigengenes.txt"), header=T, row.names=1, sep="\t")
  
  # Get submodule names and remove unclassified
  submodule.names <- unique(final.membership$Module) %>% as.character()
  submodule.names <- intersect(rownames(eigengenes), submodule.names)
  
  # Calculate submodule sizes with UNCLASSIFIED included
  submodule.sizes <- c()
  for (j in 1:length(submodule.names)) {
    submodule.sizes <- c(submodule.sizes, nrow(subset(final.membership, Module==submodule.names[j])))
  }
  submodule.names <- c(submodule.names[order(submodule.sizes, decreasing=T)], "UNCLASSIFIED")
  
  # Add information to breakdown
  for (j in 1:length(submodule.names)) {
    
    breakdown[[length(breakdown)+1]] <- cbind(
      Module=module.names[i],
      Genes=nrow(subset(final.membership, Module==submodule.names[j])),
      Submodule=ifelse(submodule.names[j]=="UNCLASSIFIED", "UNCLASSIFIED", j)
    )
  }
  
  # Remove UNCLASSIFIED information
  submodule.names <- submodule.names[-grep("UNCLASSIFIED", submodule.names)]
  
  # Recalculate sizes without UNCLASSIFIED
  #submodule.sizes <- c()
  #for (j in 1:length(submodule.names)) {
  #  submodule.sizes <- c(submodule.sizes, nrow(subset(final.membership, Module==submodule.names[j])))
  #}
  
  # Create new submodule names
  #submodule.names <- submodule.names[order(submodule.sizes, decreasing=T)]
  submodule.names.col <- paste0(module.names[i], "_", 1:length(submodule.sizes))
  
  # Save genes
  for (j in 1:length(submodule.names)) {
    gene.set[[submodule.names.col[j]]] <- subset(final.membership, Module==submodule.names[j]) %>% 
      dplyr::select(., Submodule=Module, Gene=Gene, Value=kME) 
    gene.set[[submodule.names.col[j]]]$Submodule <- submodule.names.col[j]
  }
  
  # Save eigengenes
  for (j in 1:length(submodule.names)) {
    eigengene.set[[submodule.names.col[j]]] <- t(eigengenes[submodule.names[j],])
    colnames(eigengene.set[[submodule.names.col[j]]])[1] <- submodule.names.col[j]
  }
  
  # Save submodule labels
  submodule.labels <- append(submodule.labels, as.list(submodule.names.col))
}

# Create data frame to store significant genes from each submodule
submodule.sig.genes <- do.call(rbind, gene.set)
rownames(submodule.sig.genes) <- NULL

# Create data frame to store eigengenes of cohort
eigengenes <- data.frame(Patient=cleaned$Patient, Diagnosis=cleaned$Diagnosis, Source.Diagnosis=cleaned$Source.Diagnosis)
eigengenes <- cbind(eigengenes, do.call(cbind, eigengene.set))
eigengenes$Patient <- as.character(eigengenes$Patient)
rownames(eigengenes) <- eigengenes$Patient

saveRDS(eigengenes, file=paste0("clean_data/3_iterative_WGCNA_cleaning/rosmap_eigengenes.RDS"))
saveRDS(submodule.labels, file=paste0("clean_data/3_iterative_WGCNA_cleaning/rosmap_submodule_labels.RDS"))
saveRDS(submodule.sig.genes, file=paste0("clean_data/3_iterative_WGCNA_cleaning/rosmap_submodule_sig_genes.RDS"))

# Build data frame for breakdown graph
breakdown <- do.call("rbind", breakdown) %>% as.data.frame()
breakdown$Genes <- as.numeric(levels(breakdown$Genes))[breakdown$Genes]
```

I generated graphs to see how `iterativeWGCNA` decided to split Ben's modules. I ran the program on each module invidiually and every submodule has a mutually exclusive set of genes. If genes did not highly correlate in any given module, they were unclassified by the end of the `iterativeWGCNA` process and are not included in any submodule.

```{r}
# Analyze the breakdown of modules into submodules
print(
  ggplot(data=breakdown) + 
    geom_bar(aes(x=Module, y=Genes, fill=Submodule, color=Submodule), stat="identity") +
    scale_fill_hue(l=60, c=100) +
    scale_color_hue(l=10, c=10) +
    ggtitle("Breakdown of Modules into Submodules using Iterative WGCNA") +
    basic_theme_big
)
 
ggsave(filename=paste0("results/3_iterative_WGCNA_cleaning/rosmap_iterative_wgcna_breakdown.png"), width=6, height=5)
```

