---
title: "Stratification"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)
library(lme4)
library(NbClust)
library(clusterProfiler)
library(org.Hs.eg.db)
library(cluster)
library(biwt)

source("../utils/stratification_helpers.R")
source("../utils/plot_theme.R")
```

```{r load_data}
eigengenes <- readRDS("clean_data/3_iterative_WGCNA_cleaning/rosmap_eigengenes.RDS")

module.names <- readRDS("clean_data/3_iterative_WGCNA_cleaning/rosmap_submodule_labels.RDS")
module.names <- unlist(module.names)

sig.genes <- readRDS("clean_data/3_iterative_WGCNA_cleaning/rosmap_submodule_sig_genes.RDS")

cleaned <- readRDS("clean_data/1_rosmap_cleaning/rosmap_cleaned.RDS")

# Separate into cases and controls
eigengenes.cases <- eigengenes[eigengenes$Diagnosis=="AD",]
eigengenes.controls <- eigengenes[eigengenes$Diagnosis=="CONTROL",]
```

Generate Matrix of Submodule Eigengenes
---------------------------------------

Here I created a matrix of eigengene expression data. I created separate matrices for cases and controls as well as a combined expression matrix.

```{r}
# Eigengene expression of cases
module.vecs.cases <- eigengenes.cases[,module.names]
# Eigengene expression of controls and MCIs
module.vecs.controls <- eigengenes.controls[,module.names]
# Total eigengene expression
module.vecs <- eigengenes[,module.names]
```

Cluster Analysis
----------------

I performed cluster analysis on the eigengene expression data. I used the `NbClust` package, which uses democratized results across 30 different mathematical indices to determine the optimal number of clusters. I tested one reallocation method - K-Means - and three agglomerative methods - Ward D2, McQuitty (WPGMA), and Average (UPGMA). I used silhouettes as a method of determining the optimal clustering strategy. I used Tukey's Biweight to determine relations between patients rather than Pearson correlations or Euclidean distances since Tukey's Biweight is considered a robust metric.

```{r}
# Generate a distance matrix based on Tukey's Biweight
tukey.dist <- dist(biwt.cor(module.vecs.cases)) 

# Methods tested
methods <- c("kmeans", "average", "mcquitty", "ward.D2")
nbc.res <- list()
silhouettes <- list()

set.seed(23498239)

for (m in methods) {
  
  nbc.res[[m]] <- NbClust(
    module.vecs.cases,
    diss=tukey.dist,
    distance=NULL,
    min.nc=2,
    max.nc=15,
    method=m,
    index="all"
  )
  
  s <- silhouette(nbc.res[[m]]$Best.partition, tukey.dist)
  s.m <- data.frame(matrix(as.vector(s), nrow=nrow(s)))
  silhouettes[[m]] <- s.m
  colnames(s.m) <- dimnames(s)[[2]]
  s.m <- s.m %>% 
    dplyr::arrange(cluster, sil_width) %>%
    dplyr::mutate(ID=1:nrow(s))
  s.m$cluster <- as.factor(s.m$cluster)
  print(
    ggplot(s.m) +
      geom_bar(aes(x=ID, y=sil_width, fill=cluster), stat="identity", width=1) +
      scale_fill_brewer(palette="Dark2") +
      labs(x="", y="Silhouette Width", fill="Cluster") +
      coord_flip() +
      ylim(-0.5, 0.8) +
      annotate("text", x=151, y=-0.3, label=paste0("Mean Silhouette Width: ", round(mean(s.m$sil_width), 4))) +
      basic_theme_big +
      theme(axis.text.y=element_blank(), axis.ticks.y=element_blank())
  )
  ggsave(paste0("results/4_stratification/silhouette_", m, ".svg"), width=8, height=8)
}
```

Looking at the results, I decided to use K-means since it has the best silhouette results. All four methods predicted 2 clusters.

```{r}
# Choose K-Means for the rest of the analysis
nbc.res <- nbc.res[["kmeans"]]

# Create a patient to subtype map
subtypes <- data.frame(
  Patient=rownames(module.vecs),
  Subtype=c("A", "B")[as.numeric(nbc.res$Best.partition[rownames(module.vecs)])],
  stringsAsFactors=F
)
subtypes$Subtype[is.na(subtypes$Subtype)] = "Control"
subtypes$Subtype <- factor(subtypes$Subtype, levels=c("A", "B", "Control"))
rownames(subtypes) <- subtypes$Patient

saveRDS(nbc.res, "clean_data/4_stratification/rosmap_clustering.RDS")
saveRDS(subtypes, "clean_data/4_stratification/rosmap_subtypes_map.RDS")

# Run PCA to splay our data
pca <- prcomp(module.vecs.cases)
plot.data <- as.data.frame(pca$x[,1:2])
plot.data$Subtype <- subtypes[rownames(plot.data),"Subtype"]

ggplot(data=plot.data) +
  geom_point(aes(PC1, PC2, color=Subtype), size=I(3), alpha=0.6) +
  guides(color=guide_legend(title="Subtype")) +
  scale_color_brewer(palette="Dark2") +
  guides(size=FALSE) +
  basic_theme_big
ggsave(filename="results/4_stratification/rosmap_clustering.svg", width=8, height=7.5)
```

Stratification Matrix
---------------------

I use a heatmap-like matrix to demonstrate how the eigengenes associate with specific subtypes generated in the clustering process.

```{r}
strat.mtx <- do.call(rbind, 
  lapply(
    split(module.vecs[subtypes$Patient,], subtypes$Subtype),
    function(x) apply(x, 2, mean)
  )
)
strat.mtx <- strat.mtx[nrow(strat.mtx):1,]
strat.mtx <- t(apply(strat.mtx, 1, function(x) x / apply(abs(strat.mtx), 2, max)))

p <- plot.strat(strat.mtx, main="", xlab="", ylab="", text=F)
p$layers <- p$layers[1]
p <- p + guides(fill=F) +
  theme(
    axis.ticks.x=element_blank(), 
    axis.ticks.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank()
  )
print(p)
ggsave(file="results/4_stratification/subtypes_heatmap.svg", plot=p, width=12, height=5)
```

I also generated a heatmap based on the Reactome and Kegg annotations of individual submodules. Annotations with more than one submodule were combined using SVD. 

```{r}
eigens <- annot.eigens(eigengenes)

heat.mtx <- do.call(
  rbind, 
  lapply(
    split(eigens[subtypes$Patient,], subtypes$Subtype), 
    function(x) apply(x, 2, mean)
  )
)
heat.mtx <- heat.mtx[nrow(heat.mtx):1,]
heat.mtx <- t(apply(heat.mtx, 1, function(x) x / apply(abs(heat.mtx), 2, max)))

p <- plot.strat(heat.mtx, main="", xlab="", ylab="", text=F)
p$layers <- p$layers[1]
p <- p + guides(fill=F) +
  theme(
    axis.ticks.x=element_blank(), 
    axis.ticks.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank()
  )
print(p)
```

I used cell type specific markers from McKenzie et al. to identify submodules that contain cell type specific signatures.

```{r}
eigens <- cell.eigens(eigengenes)

heat.mtx <- do.call(
  rbind, 
  lapply(
    split(eigens[subtypes$Patient,], subtypes$Subtype), 
    function(x) apply(x, 2, mean)
  )
)
heat.mtx <- heat.mtx[nrow(heat.mtx):1,]
heat.mtx <- t(apply(heat.mtx, 1, function(x) x / apply(abs(heat.mtx), 2, max)))

p <- plot.strat(heat.mtx, main="", xlab="", ylab="", text=F)
p$layers <- p$layers[1]
p <- p + guides(fill=F) +
  theme(
    axis.ticks.x=element_blank(), 
    axis.ticks.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank()
  )
print(p)
```

Finally, I generated a heatmap to compare controls, MCIs, and cases.

```{r}
strat.mtx <- do.call(rbind, 
  lapply(
    split(module.vecs[cleaned$Patient,], cleaned$Tissue.Diagnosis),
    function(x) apply(x, 2, mean)
  )
)
strat.mtx <- strat.mtx[c("CONTROL", "OTHER", "AD"),]
strat.mtx <- t(apply(strat.mtx, 1, function(x) x / apply(abs(strat.mtx), 2, max)))

p <- plot.strat(strat.mtx, main="", xlab="", ylab="", text=F)
p$layers <- p$layers[1]
p <- p + guides(fill=F) +
  theme(
    axis.ticks.x=element_blank(), 
    axis.ticks.y=element_blank(),
    panel.grid.major=element_blank(),
    panel.border=element_blank()
  )
print(p)
#ggsave(file="results/4_stratification/subtypes_heatmap.svg", plot=p, width=12, height=5)
```

Subtype Specificity Metric
--------------------------

I generated the submodule specificity metric here. I save it for future use in R as well as files that follow PLINK standards to run through EMMAX.

```{r}
subs <- as.character(unique(subtypes$Subtype[subtypes$Subtype %in% LETTERS]))
subs <- subs[ordered(subs)]

centroids <- lapply(
  split(module.vecs[subtypes$Patient,], subtypes$Subtype)[subs],
  function(x) { apply(x, 2, mean) }
)
patient.subtypes <- as.data.frame(cbind(
  rownames(module.vecs),
  as.character(subtypes[rownames(module.vecs),"Subtype"]),
  do.call(
    cbind,
    lapply(
      centroids,
      function(x) module.vecs %>%
        apply(., 1, function(v) sqrt(sum((x-v)^2)))
    )
  )
))
colnames(patient.subtypes) <- c("Patient", "Subtype", "A", "B")
patient.subtypes$A <- as.numeric(as.character(patient.subtypes$A))
patient.subtypes$B <- as.numeric(as.character(patient.subtypes$B))

saveRDS(patient.subtypes, "clean_data/4_stratification/rosmap_patient_subtypes.RDS")
```

Covariate Analysis
------------------

I tested if any covariates were enriched in any given subtype and may explain the variation in gene expression that led to the subtyping.

```{r}
# Sex
trait <- factor(c("Female", "Male")[cleaned[rownames(subtypes),"Sex"]], levels=c("Female", "Male"))
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_bar(aes(x=Subtype, fill=Trait), position=position_fill(), color="black") +
  scale_fill_brewer(palette="Set3") +
  labs(x="", y="Proportion", fill="Sex") + 
  basic_theme_big +
  theme(legend.position="bottom")
ggsave(filename="results/4_stratification/trait_sex_by_subtype.svg", width=7, height=7)

# APOE4 Status
trait <- factor(c("None", "Heterozygous", "Homozygous")[cleaned[rownames(subtypes),"apoe_genotype"]+1])
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_bar(aes(x=Subtype, fill=Trait), position=position_fill(), color="black") +
  scale_fill_brewer(palette="Set3") +
  labs(x="", y="Proportion", fill="APOE4 Status") + 
  basic_theme_big +
  theme(legend.position="bottom")
ggsave(filename="results/4_stratification/trait_apoe4_by_subtype.svg", width=7, height=7)

# Education
trait <- cleaned[rownames(subtypes), "educ"]
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Years of Education") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_education_by_subtype.svg", width=7, height=7)

# Prepare Synapse Covariates
syn.clin <- read.csv("raw_data/Synapse_Rosmap_Clinical_Data.csv", header=T, stringsAsFactors=F)
key.map <- read.csv("raw_data/ROSMAP_IDkey.csv", header=T, stringsAsFactors=F)
key.map <- key.map %>% dplyr::select(projid, rnaseq_id) %>% unique
clinical.vars <- merge(syn.clin, key.map, all.x=T, by="projid") %>%
  dplyr::filter(rnaseq_id != "")
rownames(clinical.vars) <- clinical.vars$rnaseq_id

# Ages of different things
trait <- clinical.vars[rownames(subtypes), "age_at_visit_max"]
trait[trait == "90+"] = "90" # Age-censored, assume just 90 for now
trait <- as.numeric(trait)
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Age at Last Visit") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_age_at_last_visit.svg", width=7, height=7)

trait <- clinical.vars[rownames(subtypes), "age_death"]
trait[trait == "90+"] = "90" # Age-censored, assume just 90 for now
trait <- as.numeric(trait)
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Age at First Diagnosis") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_age_at_first_dx.svg", width=7, height=7)

# Braak Scores
trait <- factor(clinical.vars[rownames(subtypes), "braaksc"], levels=1:6)
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_bar(aes(x=Subtype, fill=Trait), position=position_fill(), color="black") +
  scale_fill_brewer(palette="Set3") +
  labs(x="", y="Proportion", fill="Braak Stage") + 
  basic_theme_big +
  theme(legend.position="bottom")
ggsave(filename="results/4_stratification/trait_braak_stage.svg", width=7, height=7)

trait <- factor(clinical.vars[rownames(patient.subtypes), "braaksc"], levels=1:6)
plot.data <- patient.subtypes
plot.data$Trait <- trait
plot.data <- reshape2::melt(plot.data, id.vars=c(1,2,5))
colnames(plot.data)[4:5] <- c("SubtypeProfile", "Distance")
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Trait, y=log10(Distance), fill=SubtypeProfile)) +
  guides(fill=F) +
  labs(x="Braak Stage", y="log10(Distance from Subtype)") +
  basic_theme_big
ggsave(filename="results/4_stratification/distance_braak_stage.svg", width=7, height=7)

# CERAD Scores
trait <- factor(clinical.vars[rownames(subtypes), "ceradsc"], levels=1:4)
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_bar(aes(x=Subtype, fill=Trait), position=position_fill(), color="black") +
  scale_fill_brewer(palette="Set3") +
  labs(x="", y="Proportion", fill="CERAD Stage") + 
  basic_theme_big +
  theme(legend.position="bottom")
ggsave(filename="results/4_stratification/trait_cerad_score.svg", width=7, height=7)

trait <- factor(clinical.vars[rownames(patient.subtypes), "ceradsc"], levels=1:4)
plot.data <- patient.subtypes
plot.data$Trait <- trait
plot.data <- reshape2::melt(plot.data, id.vars=c(1,2,5))
colnames(plot.data)[4:5] <- c("SubtypeProfile", "Distance")
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Trait, y=log10(Distance), fill=SubtypeProfile)) +
  guides(fill=F) +
  labs(x="CERAD Score", y="log10(Distance from Subtype)") +
  basic_theme_big
ggsave(filename="results/4_stratification/distance_cerad_score.svg", width=7, height=7)

# Cognitive Diagnosis
trait <- factor(clinical.vars[rownames(subtypes), "cogdx"], levels=1:6)
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_bar(aes(x=Subtype, fill=Trait), position=position_fill(), color="black") +
  scale_fill_brewer(palette="Set3") +
  labs(x="", y="Proportion", fill="Cognitive Diagnosis") + 
  basic_theme_big +
  theme(legend.position="bottom")
ggsave(filename="results/4_stratification/trait_cogdx.svg", width=7, height=7)

trait <- factor(clinical.vars[rownames(patient.subtypes), "cogdx"], levels=1:6)
plot.data <- patient.subtypes
plot.data$Trait <- trait
plot.data <- reshape2::melt(plot.data, id.vars=c(1,2,5))
colnames(plot.data)[4:5] <- c("SubtypeProfile", "Distance")
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Trait, y=log10(Distance), fill=SubtypeProfile)) +
  guides(fill=F) +
  labs(x="CERAD Score", y="log10(Distance from Subtype)") +
  basic_theme_big
ggsave(filename="results/4_stratification/distance_cogdx.svg", width=7, height=7)

# Prepare Joey's data
joey.clin <- read.delim("../raw_data/Joey_Rosmap_Cognitive_Scores_Slopes.txt", header=T, stringsAsFactors=F)
key.map <- read.csv("raw_data/ROSMAP_IDkey.csv", header=T, stringsAsFactors=F)
key.map <- key.map %>% dplyr::select(projid, rnaseq_id) %>% unique
clinical.vars <- merge(joey.clin, key.map, all.x=T, by="projid") %>%
  dplyr::filter(rnaseq_id != "")
rownames(clinical.vars) <- clinical.vars$rnaseq_id

# Memory Decline
trait <- clinical.vars[rownames(subtypes), "slope_nmem"]
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Rate of Memory Decline") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_memory_decline.svg", width=7, height=7)
t1 <- plot.data[plot.data$Subtype == "A", "Trait"]
t2 <- plot.data[plot.data$Subtype == "B", "Trait"]
t.test(t1, t2)

# Visuospatial Decline
trait <- clinical.vars[rownames(subtypes), "slope_nvsp"]
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Rate of Visuospatial Decline") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_visuospatial_decline.svg", width=7, height=7)
t1 <- plot.data[plot.data$Subtype == "A", "Trait"]
t2 <- plot.data[plot.data$Subtype == "B", "Trait"]
t.test(t1, t2)

# Executive Function Decline
trait <- clinical.vars[rownames(subtypes), "slope_nexf"]
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Rate of Executive Function Decline") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_executive_function_decline.svg", width=7, height=7)
t1 <- plot.data[plot.data$Subtype == "A", "Trait"]
t2 <- plot.data[plot.data$Subtype == "B", "Trait"]
t.test(t1, t2)

# Language Decline
trait <- clinical.vars[rownames(subtypes), "slope_nlan"]
plot.data <- subtypes
plot.data$Trait <- trait
ggplot(data=na.omit(plot.data)) +
  geom_boxplot(aes(x=Subtype, y=Trait, fill=Subtype)) +
  scale_fill_brewer(palette="Set3") +
  guides(fill=F) +
  labs(x="", y="Rate of Language Decline") +
  basic_theme_big
ggsave(filename="results/4_stratification/trait_language_decline.svg", width=7, height=7)
t1 <- plot.data[plot.data$Subtype == "A", "Trait"]
t2 <- plot.data[plot.data$Subtype == "B", "Trait"]
t.test(t1, t2)
```

Generating Phenotypes for EMMAX
-------------------------------

I generated the `.pheno` file required specifically by EMMAX.

```{r}
# Load mapping from RNA Seq IDs to WGS IDs
rosmap.ids <- read.csv("raw_data/ROSMAP_IDkey.csv", stringsAsFactors=F)

# Load EMMAX .tfam file and order data by expected patients
tfam <- read.table("raw_data/Rosmap_allSNPjointMAF05_Clean__TRANSPOSED_EMMAX.tfam")
colnames(tfam) <- c("Family", "WGS.ID", "X1", "X2", "X3", "X4")

# Use cases, controls, and MCIs
subs <- subtypes

# Merge data into subtype map by RNA Seq IDs
emmax.pheno <- merge(subs, rosmap.ids, by.x="Patient", by.y="rnaseq_id") %>% 
  dplyr::select(Patient, WGS.ID=wgs_id) %>% 
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into patient subtype specificity metric data
emmax.pheno <- merge(emmax.pheno, patient.subtypes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, WGS.ID, A, B) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into eigengenes
emmax.pheno <- merge(emmax.pheno, eigengenes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, WGS.ID, A, B, module.names) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Remove any patients without WGS ID's (were not sequenced)
mask <- emmax.pheno$WGS.ID != ""& !is.na(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno[mask,]

# Align to TFAM file
emmax.pheno <- merge(tfam, emmax.pheno, by.x="WGS.ID", by.y="WGS.ID", all.x=T) %>%
  dplyr::select(c("WGS.ID", "A", "B", module.names))
rownames(emmax.pheno) <- emmax.pheno$WGS.ID
emmax.pheno <- emmax.pheno[tfam$WGS.ID,]

# Create EMMAX format file
# 1. Patient Family ID (same as patient ID since unrelated)
# 2. Patient ID
# 3. Phenotype
wgs.id <- as.character(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno %>%
  dplyr::select(c("A", "B", module.names))
for (pheno in colnames(emmax.pheno)) {
  write.table(
    cbind(wgs.id, wgs.id, emmax.pheno[,pheno]),
    file=paste0("clean_data/4_stratification/emmax_phenotypes/", pheno, ".pheno"),
    quote=F, sep="\t", row.names=F, col.names=F
  )
}
```

EMMAX files were also generated containing only MCI and Case patients. 

```{r}
# Load mapping from RNA Seq IDs to WGS IDs
rosmap.ids <- read.csv("raw_data/ROSMAP_IDkey.csv", stringsAsFactors=F)

# Load EMMAX .tfam file and order data by expected patients
tfam <- read.table("raw_data/Rosmap_allSNPjointMAF05_Clean__TRANSPOSED_EMMAX.tfam")
colnames(tfam) <- c("Family", "WGS.ID", "X1", "X2", "X3", "X4")

# Use only cases and MCIs
subs <- subtypes[cleaned[subtypes$Patient,"Tissue.Diagnosis"] != "CONTROL",]

# Merge data into subtype map by RNA Seq IDs
emmax.pheno <- merge(subs, rosmap.ids, by.x="Patient", by.y="rnaseq_id") %>% 
  dplyr::select(Patient, Subtype, WGS.ID=wgs_id) %>% 
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into patient subtype specificity metric data
emmax.pheno <- merge(emmax.pheno, patient.subtypes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, Subtype=Subtype.x, WGS.ID, A, B) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into eigengenes
emmax.pheno <- merge(emmax.pheno, eigengenes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, Subtype, WGS.ID, A, B, module.names) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Remove any patients without WGS ID's (were not sequenced)
mask <- emmax.pheno$WGS.ID != ""& !is.na(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno[mask,]

# Align to TFAM file
emmax.pheno <- merge(tfam, emmax.pheno, by.x="WGS.ID", by.y="WGS.ID", all.x=T) %>%
  dplyr::select(c("WGS.ID", "A", "B", module.names))
rownames(emmax.pheno) <- emmax.pheno$WGS.ID
emmax.pheno <- emmax.pheno[tfam$WGS.ID,]

# Create EMMAX format file
# 1. Patient Family ID (same as patient ID since unrelated)
# 2. Patient ID
# 3. Phenotype
wgs.id <- as.character(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno %>%
  dplyr::select(c("A", "B", module.names))
for (pheno in colnames(emmax.pheno)) {
  write.table(
    cbind(wgs.id, wgs.id, emmax.pheno[,pheno]),
    file=paste0("clean_data/4_stratification/emmax_phenotypes/", pheno, "_Case_MCI.pheno"),
    quote=F, sep="\t", row.names=F, col.names=F
  )
}
```

EMMAX files were also created for only Cases. 

```{r}
# Load mapping from RNA Seq IDs to WGS IDs
rosmap.ids <- read.csv("raw_data/ROSMAP_IDkey.csv", stringsAsFactors=F)

# Load EMMAX .tfam file and order data by expected patients
tfam <- read.table("raw_data/Rosmap_allSNPjointMAF05_Clean__TRANSPOSED_EMMAX.tfam")
colnames(tfam) <- c("Family", "WGS.ID", "X1", "X2", "X3", "X4")

# Use only cases
subs <- subtypes[cleaned[subtypes$Patient,"Tissue.Diagnosis"] == "AD",]

# Merge data into subtype map by RNA Seq IDs
emmax.pheno <- merge(subs, rosmap.ids, by.x="Patient", by.y="rnaseq_id") %>% 
  dplyr::select(Patient, Subtype, WGS.ID=wgs_id) %>% 
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into patient subtype specificity metric data
emmax.pheno <- merge(emmax.pheno, patient.subtypes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, Subtype=Subtype.x, WGS.ID, A, B) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Merge data into eigengenes
emmax.pheno <- merge(emmax.pheno, eigengenes, by.x="Patient", by.y="Patient") %>%
  dplyr::select(Patient, Subtype, WGS.ID, A, B, module.names) %>%
  dplyr::filter(Patient %in% subs$Patient) %>%
  unique

# Remove any patients without WGS ID's (were not sequenced)
mask <- emmax.pheno$WGS.ID != ""& !is.na(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno[mask,]

# Align to TFAM file
emmax.pheno <- merge(tfam, emmax.pheno, by.x="WGS.ID", by.y="WGS.ID", all.x=T) %>%
  dplyr::select(c("WGS.ID", "A", "B", module.names))
rownames(emmax.pheno) <- emmax.pheno$WGS.ID
emmax.pheno <- emmax.pheno[tfam$WGS.ID,]

# Create EMMAX format file
# 1. Patient Family ID (same as patient ID since unrelated)
# 2. Patient ID
# 3. Phenotype
wgs.id <- as.character(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno %>%
  dplyr::select(c("A", "B", module.names))
for (pheno in colnames(emmax.pheno)) {
  write.table(
    cbind(wgs.id, wgs.id, emmax.pheno[,pheno]),
    file=paste0("clean_data/4_stratification/emmax_phenotypes/", pheno, "_Case.pheno"),
    quote=F, sep="\t", row.names=F, col.names=F
  )
}
```

I edit the covariate file to include only Sex as a covariate

```{r}
covar.file <- read.table("raw_data/Rosmap_All_MAF05_TRANSPOSED_EMMAX_Covariates_Case_Control.covar")
covar.file <- covar.file %>% dplyr::select(V1, V2, V3, V5)
write.table(
  covar.file, "clean_data/4_stratification/emmax_phenotypes/Rosmap_Covars.covar", 
  quote=F, sep="\t", row.names=F, col.names=F
)
```

Generate EMMAX Files for Hierarchical Genetic Analysis
------------------------------------------------------

I will be running four analyses using EMMAX:
1. Case-Control Association
2. Subtype Association
3. Module Association
4. Submodule Association
5. CERAD/BRAAK/CogDX Association

Subtype and submodule annotation files were generated in the previous section. In this section, I will generate module eigengenes and case-control files for EMMAX. Since the point is to demonstrate the usefulness of subtyping, I am restricting the case-control analysis to only those patients that have transcriptomic data available as well.

```{r}
# Generate eigengenes for modules
modules <- read.delim("raw_data/Modules_March2018.txt", stringsAsFactors=F)
modules <- split(
  modules %>% 
    dplyr::filter(brainRegion=="DLPFC") %>%
    dplyr::select(Gene=GeneID, Module=Module), 
  modules$Module[modules$brainRegion=="DLPFC"]
)

module.eigens <- lapply(modules, function(x) {
  gene.exp = cleaned[,intersect(colnames(cleaned), x$Gene)]
  gene.exp.p = prcomp(gene.exp)
  return(gene.exp.p$x[,"PC1"])
})
names(module.eigens) <- names(modules)
module.eigens <- do.call(cbind, module.eigens)

# Generate case-control variable
# AD = 1, Control = 0
diagnosis <- as.numeric(factor(as.character(cleaned[rownames(module.eigens),"Diagnosis"]), levels=c("CONTROL", "AD"))) - 1
module.eigens <- cbind(module.eigens, diagnosis)
colnames(module.eigens)[ncol(module.eigens)] <- "Diagnosis"

# Generate case-MCI-control variable
# AD = 2, MCI = 1, Control = 0
diagnosis <- as.numeric(factor(as.character(cleaned[rownames(module.eigens),"Tissue.Diagnosis"]), levels=c("CONTROL", "OTHER", "AD"))) - 1
module.eigens <- cbind(module.eigens, diagnosis)
colnames(module.eigens)[ncol(module.eigens)] <- "Diagnosis_MCI"

# Generate CERAD/BRAAK/CogDX scores
syn.cov <- read.csv("raw_data/Synapse_Rosmap_Clinical_Data.csv", header=T)
syn.cov <- syn.cov %>% dplyr::select(projid, braaksc, ceradsc, cogdx)

# Add patients as a column
module.eigens <- cbind(rownames(module.eigens), as.data.frame(module.eigens))
colnames(module.eigens)[1] <- "Patient"
module.eigens$Patient <- as.character(module.eigens$Patient)

# Load mapping from RNA Seq IDs to WGS IDs
rosmap.ids <- read.csv("raw_data/ROSMAP_IDkey.csv", stringsAsFactors=F)

# Load EMMAX .tfam file and order data by expected patients
tfam <- read.table("raw_data/Rosmap_allSNPjointMAF05_Clean__TRANSPOSED_EMMAX.tfam")
colnames(tfam) <- c("Family", "WGS.ID", "X1", "X2", "X3", "X4")

# Add rnaseq_ids to synapse covar and merge into other phenotypes
syn.cov <- merge(rosmap.ids, syn.cov, by="projid", all.x=T) %>%
  dplyr::select(Patient=rnaseq_id, braaksc, ceradsc, cogdx) %>%
  dplyr::filter(Patient != "") %>%
  unique()
rownames(syn.cov) <- syn.cov$Patient
module.eigens$Braak.Score <- syn.cov[module.eigens$Patient,"braaksc"]
module.eigens$CERAD.Score <- syn.cov[module.eigens$Patient,"ceradsc"]
module.eigens$Cog.Diagnosis <- syn.cov[module.eigens$Patient,"cogdx"]

# Merge data into subtype map by RNA Seq IDs
emmax.pheno <- merge(module.eigens, rosmap.ids, by.x="Patient", by.y="rnaseq_id") %>% 
  dplyr::select(Patient, WGS.ID=wgs_id, colnames(module.eigens)) %>% 
  dplyr::filter(Patient %in% module.eigens$Patient) %>%
  unique

# Remove any patients without WGS ID's (were not sequenced)
mask <- emmax.pheno$WGS.ID != ""& !is.na(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno[mask,]

# Align to TFAM file
emmax.pheno <- merge(tfam, emmax.pheno, by.x="WGS.ID", by.y="WGS.ID", all.x=T) %>%
  dplyr::select(c("WGS.ID", colnames(module.eigens)))
rownames(emmax.pheno) <- emmax.pheno$WGS.ID
emmax.pheno <- emmax.pheno[tfam$WGS.ID,]

# Create EMMAX format file
# 1. Patient Family ID (same as patient ID since unrelated)
# 2. Patient ID
# 3. Phenotype
wgs.id <- as.character(emmax.pheno$WGS.ID)
emmax.pheno <- emmax.pheno %>%
  dplyr::select(colnames(module.eigens)[-1])
for (pheno in colnames(emmax.pheno)) {
  write.table(
    cbind(wgs.id, wgs.id, emmax.pheno[,pheno]),
    file=paste0("clean_data/4_stratification/emmax_phenotypes/", pheno, ".pheno"),
    quote=F, sep="\t", row.names=F, col.names=F
  )
}
```

