---
title: "MAYO Cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)
library(ggdendro)

source("../utils/plot_theme.R")
```

```{r load_data}
labels <- read.table("raw_data/MAYO_CBE_TCX_Covariates.tsv", header=TRUE, sep="\t", row.names=1)
cleaned <- read.table("raw_data/MAYO_CBE_TCX_netResidualExpression_imputed.tsv", header=TRUE, sep="\t", row.names=1)

# Fix column names
colnames(cleaned) <- substring(colnames(cleaned), 2)

# Extract the patient IDs and gene names
patient.key <- colnames(cleaned)

# Match data structures
cleaned <- t(cleaned)
cleaned <- apply(cleaned, 2, as.numeric)
cleaned <- as.data.frame(cleaned)

# Merge labels (covariates) and data
cleaned$Patient <- patient.key
labels$Patient <- row.names(labels)
cleaned <- merge(labels, cleaned, by="Patient")
```

Principle Component Analysis
----------------------------

The data has been extensively transformed. We will perform PCA decomposition on the expression data to make sure that there are no obvious outliers.

```{r}
# Any resulting PCs with SD < 0.01 are not included
pca.results <- cleaned[,(ncol(labels)+1):ncol(cleaned)] %>%
  prcomp(scale=F, center=T, tol=0.01)
summary(pca.results)$importance[1:2]

# Visualize the PCA
pca.plot <- cbind(Diagnosis=cleaned$Tissue.Diagnosis, as.data.frame(pca.results$x[,1:2]))
ggplot(data=pca.plot, aes(x=PC1, y=PC2, color=Diagnosis)) +
  geom_point(size=I(2), alpha=0.5) +
  ggtitle("PCA of Samples Colored by Diagnosis") +
  xlab("PC1 (21.6713%)") + 
  ylab("PC2 (20.0321%)") +
  basic_theme_big
ggsave("results/1_mayo_cleaning/mayo_diagnosis_PCA.png", width=8, height=8)
```

Hierarchical Clustering
-----------------------

Generate a dendrogram to visualize relationships between samples.

```{r}
# Create dendrogram
row.names(cleaned) <- 1:nrow(cleaned)
cleaned$ID <- 1:nrow(cleaned)
dend.expr <- cleaned[,(ncol(labels)+1):ncol(cleaned)] %>%
  scale %>%
  dist %>%
  hclust %>%
  as.dendrogram()

# Plot dendrogram
dendro.data <- dendro_data(dend.expr, type="rectangle")
dendro.data$labels <- merge(x=dendro.data$labels, y=cleaned, by.x="label", by.y="ID")
dendro.data$labels[,"label"] <- patient.key[dendro.data$labels[,"label"]]

# Plot dendrogram with colored branches and legend
ggplot() +
  geom_segment(data=segment(dendro.data), aes(x=x, y=y, xend=xend, yend=yend)) +
  geom_segment(data = dendro.data$segments %>%
                 filter(yend == 0) %>%
                 left_join(dendro.data$labels, by="x"), aes(x=x, y=y.x, xend=xend, yend=yend, color=Tissue.Diagnosis)) +
  geom_text(data=label(dendro.data), aes(x=x, y=y, label=label, color=Tissue.Diagnosis, hjust=0), size=1) +
  coord_flip() +
  scale_y_reverse(expand=c(0.2, 0)) +
  scale_colour_brewer(palette="RdBu") + 
  basic_theme_big +
  theme_dendro() +
  ggtitle("Hierarchical Clustering of Patients Colored by Diagnosis")
ggsave("results/1_mayo_cleaning/mayo_diagnosis_hierarchial_clustering.png", width=8, height=8)
```

Tissue Removal and Cleaning
---------------------------

Here, we normalize the data across the cohorts. We remove the CBE data from MAYO and choose to focus on the TCX tissue. We retain the source diagnosis in a separate column but choose to treat PSP and Pathological Aging diagnoses as controls for downstream analyses. 

```{r}
# Remove ID column created previously
cleaned <- cleaned[,-which(colnames(cleaned) == "ID")]

# Remove CBE Data
cleaned <- cleaned[grep("TCX", cleaned$Patient),]
patient.key <- cleaned$Patient[grep("TCX", cleaned$Patient)]

# Remove tissue name from diagnosis
src.diagnosis <- gsub("TCX.", "", cleaned$Tissue.SourceDiagnosis)
diagnosis <- src.diagnosis

# Convert Progressive Supranuclear Palsy and Pathological Age to Controls
diagnosis <- gsub("PSP", "CONTROL", diagnosis)
diagnosis <- gsub("PATH_AGE", "CONTROL", diagnosis)

# Add diagnosis column before genes
cleaned <- cbind(cleaned[,1:18], diagnosis, src.diagnosis, cleaned[,-(1:18)])
colnames(cleaned)[19] <- "Diagnosis"
colnames(cleaned)[20] <- "Source.Diagnosis"

# Get final list of genes
gene.key <- colnames(cleaned)[-(1:20)]

# Set patients as key to cleaned data
rownames(cleaned) <- cleaned$Patient
```

```{r}
# Save data for further analysis
saveRDS(patient.key, file=paste0("clean_data/1_mayo_cleaning/mayo_patient_key.RDS"))
saveRDS(gene.key, file=paste0("clean_data/1_mayo_cleaning/mayo_gene_key.RDS"))
saveRDS(cleaned, file=paste0("clean_data/1_mayo_cleaning/mayo_cleaned.RDS"))
```

