---
title: "MSSM Cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggdendro)
library(ggplot2)

source("../utils/plot_theme.R")
```

```{r load_data}
labels <- read.table("raw_data/MSSM_ageCensoredCovariates.tsv", header=TRUE, sep="\t", row.names=1)
metadata <- read.csv("raw_data/MSSM_metadata.csv", header=TRUE)
rnaseq.covars <- read.csv("raw_data/MSSM_RNASeq_covars.csv", header=TRUE)
cleaned <- read.table("raw_data/MSSM_FP_STG_PHG_IFG_netResidualExpression_imputed.tsv", header=TRUE, sep="\t", row.names=1)

# Extract the patient IDs
patient.key <- colnames(cleaned)

# Match data structures
cleaned <- t(cleaned)
cleaned <- apply(cleaned, 2, as.numeric)
cleaned <- as.data.frame(cleaned)

# Merge labels (covariates) and data
cleaned$Patient <- patient.key
labels$Patient <- row.names(labels)

# Add Amyloid Plaque
id.match <- unique(rnaseq.covars[,c(2,7)])
rownames(id.match) <- id.match$sampleIdentifier
rownames(metadata) <- metadata$RNAseq_SampleID
indiv.identifiers <- id.match[patient.key,]$individualIdentifier
cleaned <- cbind(metadata[indiv.identifiers,]$PlaqueMean, cleaned)
colnames(cleaned)[1] <- "PlaqueMean"

cleaned <- merge(labels, cleaned, by="Patient")
```

Principle Component Analysis
----------------------------

The data has been extensively transformed. We will perform PCA decomposition on the expression data to make sure that there are no obvious outliers.

```{r}
# Any resulting PCs with SD < 0.01 are not included
pca.results <- cleaned[,(ncol(labels)+1):ncol(cleaned)] %>%
  prcomp(scale=F, center=T, tol=0.01)
#summary(pca.results)$importance

# Visualize the PCA
pca.plot <- cbind(Diagnosis=as.factor(cleaned$BrainRegion.Diagnosis), as.data.frame(pca.results$x[,1:2]))
ggplot(data=pca.plot, aes(x=PC1, y=PC2, color=Diagnosis)) +
  geom_point() +
  ggtitle("PCA of Samples Colored by Diagnosis") +
  xlab("PC1 (17.4921%)") + 
  ylab("PC2 (10.89594%)") +
  basic_theme_big
ggsave("results/1_mssm_cleaning/mssm_diagnosis_PCA.png", width=8, height=8)
```

Hierarchical Clustering
-----------------------

Generate a dendrogram to visualize relationships between samples.

```{r}
# Create dendrogram
row.names(cleaned) <- 1:nrow(cleaned)
cleaned$ID <- 1:nrow(cleaned)
dend.expr <- cleaned[,(ncol(labels)+1):ncol(cleaned)] %>% # Isolate expression
  scale %>% # Normalize
  dist %>% # Calculate distance
  hclust %>% # Cluster hierarchially
  as.dendrogram()

# Plot dendrogram
dendro.data <- dendro_data(dend.expr, type="rectangle")
dendro.data$labels <- merge(x=dendro.data$labels, y=cleaned, by.x="label", by.y="ID")
dendro.data$labels[,"label"] <- patient.key[dendro.data$labels[,"label"]]

# Plot dendrogram with colored branches and legend
ggplot() +
  geom_segment(data=segment(dendro.data), aes(x=x, y=y, xend=xend, yend=yend)) +
  geom_segment(data = dendro.data$segments %>%
                 filter(yend == 0) %>%
                 left_join(dendro.data$labels, by="x"), aes(x=x, y=y.x, xend=xend, yend=yend, color=BrainRegion.Diagnosis)) +
  geom_text(data=label(dendro.data), aes(x=x, y=y, label=label, color=BrainRegion.Diagnosis, hjust=0), size=1) +
  coord_flip() +
  scale_y_reverse(expand=c(0.2, 0)) +
  scale_colour_brewer(palette="RdBu") +
  basic_theme_big +
  theme_dendro() +
  ggtitle("Hierarchical Clustering of Patients Colored by Diagnosis")
ggsave("results/1_mssm_cleaning/mssm_diagnosis_hierarchial_clustering.png", width=8, height=8)
```

Tissue Removal and Cleaning
---------------------------

Removed tissue names from the diagnoses. For the sake of future analyses, all MCI patients were considered controls and were annotated as such in the `Diagnosis` column. The `Source.Diagnosis` column keeps track of the original diagnosis.

```{r}
# Remove ID column created previously
cleaned <- cleaned[,-which(colnames(cleaned) == "ID")]

# Rename column containing diagnosis
colnames(cleaned)[colnames(cleaned)=="BrainRegion.Diagnosis"] <- "Tissue.Diagnosis"

# Remove tissue name from diagnosis
diagnosis <- gsub("FP.", "", cleaned$Tissue.Diagnosis)
diagnosis <- gsub("STG.", "", diagnosis)
diagnosis <- gsub("PHG.", "", diagnosis)
diagnosis <- gsub("IFG.", "", diagnosis)
src.diagnosis <- diagnosis

# Convert "OTHER" phenotypes to control
diagnosis <- gsub("OTHER", "CONTROL", diagnosis)

# Add diagnosis column before genes
cleaned <- cbind(cleaned[,1:19], diagnosis, src.diagnosis, cleaned[,-(1:19)])
colnames(cleaned)[20] <- "Diagnosis"
colnames(cleaned)[21] <- "Source.Diagnosis"

# Get final list of genes
gene.key <- colnames(cleaned)[-(1:21)]

# Rename sex
cleaned <- dplyr::rename(cleaned, Sex=SEX)

# Make patients the key
rownames(cleaned) <- cleaned$Patient
```

```{r}
# Save files for future analyses
saveRDS(patient.key, file=paste0("clean_data/1_mssm_cleaning/mssm_patient_key.RDS"))
saveRDS(gene.key, file=paste0("clean_data/1_mssm_cleaning/mssm_gene_key.RDS"))
saveRDS(cleaned, file=paste0("clean_data/1_mssm_cleaning/mssm_cleaned.RDS"))
```

