---
title: "Differential Expression Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())

library(dplyr)
library(ggplot2)
library(ggrepel)
library(limma)
library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r load_data, include=FALSE}
subtypes <- readRDS("clean_data/4_stratification/mssm_patient_subtypes.RDS")
subtypes$Patient <- as.character(subtypes$Patient)
cleaned <- readRDS("clean_data/1_mssm_cleaning/mssm_cleaned.RDS")
cleaned <- cleaned %>% dplyr::filter(grepl("FP", Tissue.Diagnosis))
gene.key <- readRDS("clean_data/1_mssm_cleaning/mssm_gene_key.RDS")

# Split cleaned RNA expression into covariates and expression
covars <- cleaned[,setdiff(colnames(cleaned), gene.key)]
rownames(covars) <- cleaned$Patient
expr <- cleaned[,gene.key]
rownames(expr) <- cleaned$Patient

# Determine number of subtypes
n.subs <- length(unique(subtypes$Subtype)) - 1
```

Differential Expression Analysis
--------------------------------

I use the `limma` R package to run differential expression analysis to compare subtype expression to control expression. In this analysis, control includes both Non-AD and MCI patients. 

```{r, include=FALSE}
de.res <- list()

for (subtype in LETTERS[1:n.subs]) {
  # Reduce expression to subtype and control
  patient.key <- union(
    subtypes$Patient[subtypes$Subtype == "Control"], 
    subtypes$Patient[subtypes$Subtype == subtype]
  )
  de.expr <- as.data.frame(t(expr)[,patient.key])
  
  # Create a factor describing the patient subtype
  patient.fac <- factor(c(
    rep("Control", table(subtypes$Subtype)["Control"]), 
    rep(subtype, table(subtypes$Subtype)[subtype])
  ), levels=c("Control", subtype))
  
  # Fit with LIMMA
  fit <- lmFit(de.expr, design=model.matrix(~patient.fac))
  
  # Get top differentially expressed genes
  fit <- eBayes(fit)
  tt <- topTable(fit, number=Inf, coef=2)
  print(head(tt))
  
  # Compile LIMMA results and generate volcano plot
  de.res[[subtype]] <- data.frame(
    Gene=rownames(tt),
    Log.Fold.Change=tt$logFC,
    Log.10.P.Value=-log10(tt$adj.P.Val)
  )
}
```

I created volcano plots to visualize the results of the differential expression analysis for each subtype. I labelled genes that had an absolute log fold change of greater than 1 and an adjusted p-value less than 0.1.

```{r message=FALSE, warning=FALSE}
for (subtype in LETTERS[1:n.subs]) {
  limma.res <- de.res[[subtype]]
  limma.res$Color <- "black"
  limma.res$Color[limma.res$Log.10.P.Value > 1 & limma.res$Log.Fold.Change < 0] <- "royalblue1"
  limma.res$Color[limma.res$Log.10.P.Value > 1 & limma.res$Log.Fold.Change > 0] <- "firebrick3"
  
  # Get Gene Names
  limma.res.top <- rbind(
    limma.res[limma.res$Log.Fold.Change < -1 & limma.res$Log.10.P.Value > 1,], 
    limma.res[limma.res$Log.Fold.Change > 1 & limma.res$Log.10.P.Value > 1,]
  )
  #limma.res.top$Gene = bitr(limma.res.top$Gene, fromType="ENSEMBL", toType="SYMBOL", OrgDb=org.Hs.eg.db)$SYMBOL
  bitr.res <- bitr(limma.res.top$Gene %>% as.character, fromType="ENSEMBL", toType="SYMBOL", OrgDb=org.Hs.eg.db)
  limma.res.top <- merge(limma.res.top, bitr.res, by.x="Gene", by.y="ENSEMBL")
  
  print(ggplot() +
    geom_point(data=limma.res, aes(x=Log.Fold.Change, y=Log.10.P.Value), color=limma.res$Color, size=I(1.5), alpha=0.6) +
    geom_text_repel(data=limma.res.top, aes(x=Log.Fold.Change, y=Log.10.P.Value, label=SYMBOL)) +
    geom_vline(xintercept=c(-1,1), lty=2, color="black", alpha=0.5) +
    xlab("Log Fold Change") + ylab("-log10(p-value)") +
    xlim(-2.1, 2.1) +
    theme_linedraw() +
    theme(
      panel.grid.major=element_blank(),
      panel.grid.minor=element_blank(),
      panel.border=element_blank(),
      axis.line=element_line(color="black")
    ))
  ggsave(filename=paste0("results/7_DE_analysis/volcano_", tolower(subtype), ".svg"), width=8, height=8)
}
```

```{r, include=FALSE}
# Save results for further pathway analysis
for (subtype in LETTERS[1:n.subs]) {
  limma.res <- de.res[[subtype]] %>% dplyr::select(Gene, Log.Fold.Change, Log.P.Value=Log.10.P.Value)
  limma.res$Gene = as.character(limma.res$Gene)
  rownames(limma.res) <- limma.res$Gene
  saveRDS(limma.res, paste0("clean_data/7_DE_analysis/subtype.", subtype, ".limma.RDS"))
}
```
